Regression Cube Analysis of Cohort Study Data

### 2015-03-28 draft.tex
revised using Bernhards notes

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per cohort subject.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data is too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to pairwise assess the influence of independent features towards the target.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

Zhang et al. \cite{Zhang2014} is closest to ours regarding the application of visual analysis on cohort study data.
%%
Their work present \emph{Cohort Analysis via Visual Analytics} (\emph{CAVA}), a framework that distinguishes three major elements of cohort study data analysis: \emph{cohort data} and its manipulation using operations - \emph{views} and \emph{analytics}.
%%
They use the system to find longitudinal pathways for diseases using subject health records.
%%
Sub-groups are created automatically by applying \emph{analytics} and expert-guided using \emph{views}.
%%
Unfortunately, they do not go into the detail into the analytics used, they seem to not incorporate confounder analysis (other than dividing groups by gender) and the user study lacks concrete goals.
%%
Their data differs strongly from ours, having multiple moments for each subject, yielding a approach focused on finding pathways of interest.
%%
We use their requirements as guidelines, which incorporate \emph{flexible} and \emph{iterative analysis}, meaning that the system has to adapt to different requirements in different analysis steps.
%%
We implement this by supporting both hypothesis-based and hypothesis-free analysis.

Angelelli et al. \cite{Angelelli} is similar to ours as they visualize both image-derived an non image data using cube data structures with focus on comparison and knowledge extraction.
%%
Their proposed engine represents multiple heterogenous cohort study data sets as multiple normalized data cubes to eliminate data redundancy between them and perform runtime measure aggregation when measures of the different cubes have to be combined.
%%
This allows for saving memory space.
%%
Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{Pearsson's r} to depict relationships towards target features and employ list views and scatter plots to visualize and rank them.
%%
We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
We employ the idea of fast conducting large scale correlation analysis between many features to derive insight into the data.

\paragraph{Visual Analysis in Public Health.}
%%--
Shneiderman et al. \cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
%%
They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
%%
We focus on the latter.
%%
%Shneiderman and colleagues also present challenges, 
The presented challenges in this application comprise custom tailoring the systems to be comprehensible and intuitive to fit into the short time span clinicians have when analyzing data.
%%
Futhermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.

@@

%%
Schreck and Keim \cite{Schreck} presented a method for analyzing social media data with focus on characterization and cause of an epidemic.
%%
They employ data aggregation techniques and use geographical data to map social media events, focusing on word clouds to derive insight into possible causes.
%%
They derive hypotheses using their high variation average message density algorithms, steering the users focus on particular events.
%%
The focus remains on geolocated social media data rather than subjects with multiple dimensions.
%%
These and similar systems follow the same analysis work as we do to derive insight into data.

Perer et al. divide cohort analysis in three major parts, present \emph{CAVA}, a system which allows

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al. \cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates and time plots to display and select data.
%%
Medical image data can be incorporated to investigate potentially interesting individuals.
%%
We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.

%%
Generalized Pairs plots (\texttt{GPLOM´S}) extend the idea of scatter plot matrices by pairwise visualizing heterogeneous data using type-combination-dependent visualizations \cite{GPLOMS, Francois2013}.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter.
%%
Dai et al. \cite{Dai2005} incorporate small multiples using a \texttt{GPLOM}-like visualization using choropleth maps mapping spatial data, such as mortality rates together with scatter plots augmented with Pearsson's r values.
%%
A \emph{concept map} summarized features related to a specified disease.
%%
Time-dependent epidemiological data is visualized by Chui et al. \cite{Chui2011} using multi-panel graphs highlighting risk factor differences with age and gender towards influenza- and salmonellosis-associated hospitalizations.
%%
We also include confounder features (age, gender) into the analysis while focusing on non-time-dependent cohort study data sets.
%%

\paragraph{Statistical based Analysis.}
%%
Statistical analysis are an inherent part of Visual Analysis.
%%
Here we present work showing a strong focus on statistical analysis in this context.
%% Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization
Bertini et al. \cite{Bertini} present a overview over quality metrics describing high dimensional data.
%%
Their research agenda acts as guideline for our design, comprising application using domain-experts, perceptual tuning towards human pattern recognition of important aspects and scalability between different data set sizes.
%%
%% TODO: Use this one? "Statistical resilience of random populations to random perturbations" - http://www.ncbi.nlm.nih.gov/pubmed/19256997 - wie eine Variable sich ändert, wenn man per Zufall die Populationen verändert, um zu schauen wie zuverlässig diese Statistik ist.
%% The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems
Ahmadi et al. \cite{Ahmadi} define the \emph{Sparse Regression Cube}, which partitions sparse high-dimensional into subspaces, which are then described using by their most reliable linear regression model.
%%
In their application, they try to find most fuel-efficient roads connecting two user-defined landmarks.
%%
They focus on presenting a algebraic for efficient regression-model calculation to find the best fit for a subspace.
%%
The dimensions for the calculations are fixed, while we assess different features for each model calculation and can therefore not benefit from a hierarchical calculation approach.

%% Hierarchical Brushing of High-Dimensional Data Sets Using Quality Metrics
Albuquerque et al. \cite{Albuquerque} present a interactive exploration framework displaying quality metrics of high dimensional data sets with brushing facilities to create subsets.
%%
These subsets are then analyzed in detail, following a drilling-down approach.
%%
They incorporate scatter plot matrices (\texttt{SPLOM´S}) of quality metrics describing the features.
%%
We incorporate the idea of abstracting features for comparison.
%%
Turkay et al. \cite{Turkay} follow a similar approach by using both descriptive metrics for features as well as the features themselves and incorporate them into linked plots.
%%
The approach was applied to a cognitive aging studies, where new hypotheses could be derived.

@@

\paragraph{Prior Work.}
In recent publications, we focused on analysis on characterization of the healthy aging process of the lumbar spine.
%%
Therefore, we utilized semiautomatic detection algorithm do detect the lumbar spine shape \cite{Klemm2013VMV}.
%%
The reliable part of the presented models are represented by the shape of the lumbar spine canal, which was extracted and described using line representation.
%%

We defined a Interactive Visual Analysis workflow for image-centric cohort study data, by extending the feature selection step (see \ref{sec:Background}) with a iterative analysis loop incorporating group selection and visualization using expert-input as well as clustering methods \cite{Klemm2014VIS}.
%%
Information visualizations of cohort study features were augmented with extracted image data, yielding linked views with both image-and non-image information.
%%
The popularity of the mosaic plot depicting \emph{Cram\'{e}r's V} contingency values with our domain experts was the inspiration for this work.
%%
The representation presented in \cite{Klemm2014VIS} only highlights pairwise correlation between features without a target disease.

%% TODO: Check DLPD for correct page numbers
While the presented techniques were well received with epidemiological experts, the explanatory power towards back pain was limited, yielding a analysis based on image-derived features, such as extracting and analyzing curvature, torsion and angle of the lumbar spine \cite{Klemm2015}.
%%
These measures were analyzed using a novel decision tree quality plot towards non-image data.
%%
We concluded that the model quality is not enough to characterize back pain.
\\\\
The difference of the presented approach towards our previous and other herein presented work is twofold:
\begin{enumerate}
	\item We focus on large scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
	\item The analysis is conducted towards a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
\end{enumerate}
%%
\section{Regression Cube Analysis of Cohort Study Data}
% TODO: Do we need a preprocessing section?
%\com{Add Data Preprocessing Section.}
%%
The basic idea of our \emph{Regression Cube} is to provide an overview visualization of large cohort study data sets towards target variables.
%%
Overview visualizations of feature relationships as presented by Angelelli et al. \cite{Angelelli} are often focused on relationships between the visualized features.
%%
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
%%
In epidemiology, these relationships are also of interest, but rather w.r.t. their explanatory power towards the target feature.
%%
These target features often indicates the presence of the investigated disease.
%%
As described in Section \ref{sec:RegressionAnalysis}, regression analyses are the statistical tool of choice for analyzing these relationships.
%%
A regression model is based on expert knowledge, there is no unified rule on how to apply them to a given set of features, so they have to be applied with care.
%%
\subsection{Cube Description using Regression Formula Notation}
%%
Expert knowledge is introduced into regression analyses using the regression formulas.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%%
%In order to calculate all combinations of two to three independent features, we can apply  \emph{Regression Cube} aims to 
%%

Since we want to use the regression analyses with an overview visualization, we are are interested in all possible combinations of (two or more) independent features describing a target.
%%
We achieve this by introducing dynamic variables $X$, $Y$ and $Z$ into the regression notation.
%%
Our method replaces the dynamic variables with all features in the data set.
%%
In a data set with $n$ (e.g. 100) features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}
would yield $n^2$ (10.000) regression models, describing all possible combinations of two features in the data describing the target $Cancer$.
%%
The major advantage of this notation is that it comes natural to anyone familiar with regression analysis, because it uses the same notation, all operators can be used as before.
%%
This allows for a fast adaptation and in the epidemiological application domain.
%%
With simple adjustments to the formula, different results can be achieved:
%%
\begin{itemize}
	\item $Z \sim X + Y$ calculates all combinations of two features towards all possible target features.
	\item $Cancer \sim X + Y + BodyWeight$ includes the $BodyWeight$ feature into all regression models as feature.
	\item $Cancer \sim X + Y + Z$ calculates all combinations of three features towards the target.
\end{itemize}
%%
The problem with this approach lies in its complexity.
%%
The number of calculated regression models increases exponentially for each dynamic variable added.
%%
If we assume a data set with 100 features with the formula $Z \sim X + Y$ we calculate 1,000,000 regression models.
%%
When each regression takes about 50~ms calculation, we have to wait roughly 14 hours for the calculation to complete.
%%
Therefore, the computational complexity needs to be reduced.
%%
%\com{Regression formulas are described using standard regression formula notation by introducing variables.}
%%
%\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}

\subsection{Target-Variable-dependent Dimension Reduction}
% \com{Uli, Can you please proofread this?}
% %%
% The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
% %%
% Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
% %%
% The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for this task \cite{CFS}.
% %%
% It uses information entropy to select the features which have the most explanatory power towards the target feature.
% %%
% At the same time it tries to reduce the correlation between the independent features.
% %%
In epidemiological studies, manifold recordings lead to an abundance of features.
%%
Generally many of them exhibit low or none correlation towards the target feature.
%%
Identifying irrelevant features and excluding them from the feature space considerably reduces computational costs and contributes to presentiveness of the Regression Cube.
%%
The Correlation-based feature selection (CFS) \cite{CFS} aims to find a feature subset that maximizes the \emph{merit value} $M_F$ which is the ratio between the average feature-class and feature-feature dependencies of features in F.
%%
The dependency of a set of features utilizes the entropy-based information gain to measure the explanatory power towards the target feature.
%%
Beginning with an empty set of features $F$ the CFS algorithm iteratively adds the feature $f$ to $F$ that leads to the highest new merit-value $M_{F\cup f}$ and halts when adding any feature does not increase the merit.
%%
When for example the \emph{body weight} has a strong explanatory power towards the target, it is likely that \emph{BMI} or \emph{waist circumference} behave similar towards it.
%%
They, however, correlate strongly with each other.
%%
The CFS algorithm would then select the feature which has the largest explanatory power and discards the other dimensions.

@@

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
$R^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of $R^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the $R^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
The goal is to create an overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

@@

 The values of the current selected slice are mirrored and represented as orange data points on the slicing plane.

@@

More importantly, it allows the user to focus all mental resources on the analysis tasks, rather than wrangling with configuring the system to serve the current task or hypothesis.

@@

Therefore, our prototype consists of three components:

@@

	\item The \emph{formula editor} allows to adjust the formula towards a current hypothesis or conducting a \emph{hypothesis-free} analysis.

@@

\subsection{Regression Cube Visualization [2.SI].}
%%
The visualization and interaction with the \emph{Regression Cube} is the heart of the prototype.
%%
Results from the statistical processors are uploaded into the visualization by slice, allowing the assessment of the data as soon as parts of the calculations are finished while the rest is still in processing.

>>>>

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per cohort subject.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data is too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to assess the influence of independent features towards the target pairwise.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

Zhang et al. \cite{Zhang2014} is closest to our work regarding the application of visual analysis on cohort study data.
%%
They present \emph{Cohort Analysis via Visual Analytics} (\emph{CAVA}), a framework that distinguishes three major elements of cohort study data analysis: \emph{cohort data} and its manipulation using operations - \emph{views} and \emph{analytics}.
%%
They use the system to find longitudinal pathways for diseases using subject health records.
%%
Sub-groups are created automatically by applying \emph{analytics} and expert-guided \emph{views}.
%%
They do not go into the detail w.r.t the analytics used, they seem to not incorporate confounder analysis (other than dividing groups by gender) and the user study lacks concrete goals.
%%
Their data differs strongly from ours, having multiple moments for each subject, yielding a approach focused on finding pathways of interest.
%%
We use their requirements as guidelines, which incorporate \emph{flexible} and \emph{iterative analysis}, meaning that the system has to adapt to different requirements in different analysis steps.
%%
We implement this by supporting both hypothesis-based and hypothesis-free analysis.

Angelelli et al. \cite{Angelelli} visualize both image-derived an non image data using cube data structures with focus on comparison and knowledge extraction.
%%
Their proposed engine represents multiple heterogenous cohort study data sets as multiple normalized data cubes to eliminate data redundancy between them and perform runtime measure aggregation when measures of the different cubes have to be combined.
%%
%This allows for saving memory space.
%%
%Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{Pearsson's r} to depict relationships towards target features and employ list views and scatter plots to visualize and rank them.
%%
We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
We employ the idea of fast conducting large scale correlation analysis between many features to derive insight into the data.

\paragraph{Visual Analysis in Public Health.}
%%--
Shneiderman et al. \cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
%%
They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
%%
We focus on the latter.
%%
%Shneiderman and colleagues also present challenges, 
The presented challenges in this application comprise custom tailoring the systems to be comprehensible and intuitive to fit into the short time span clinicians have when analyzing data.
%%
Furthermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
%We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.

@@

%%
Schreck and Keim \cite{Schreck} presented a method for analyzing social media data with focus on characterization and cause of an epidemic.
%%
They employ data aggregation techniques and use geographical data to map social media events, focusing on word clouds to derive insight into possible causes.
%%
They derive hypotheses using their high variation average message density algorithms, steering the users focus on particular events.
%%
%The focus remains on geolocated social media data rather than subjects with multiple dimensions.
%%
%These and similar systems follow the same analysis work as we do to derive insight into data.

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al. \cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates and time plots to display and select data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.

%%
Generalized Pairs plots (\texttt{GPLOM´S}) extend the idea of scatter plot matrices by pairwise visualizing heterogeneous data using type-combination-dependent visualizations \cite{GPLOMS, Francois2013}.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter and are therefore suitable for small epidemiological data sets.
%%
Dai et al. \cite{Dai2005} incorporate small multiples using a \texttt{GPLOM}-like visualization using choropleth maps mapping spatial data, such as mortality rates together with scatter plots augmented with Pearsson's r values.
%%
A \emph{concept map} summarized features related to a specified disease.
%%
Time-dependent epidemiological data is visualized by Chui et al. \cite{Chui2011} using multi-panel graphs highlighting risk factor differences with age and gender towards influenza- and salmonellosis-associated hospitalizations.
%%
We also include confounder features (age, gender) into the analysis while focusing on non-time-dependent cohort study data sets.
%%

\paragraph{Statistical Analysis.}
%%
Statistical analysis is an inherent part of Visual Analysis.
%%
Here we present work showing a strong focus on statistical analysis in this context.
%% Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization
Bertini et al. \cite{Bertini} present a overview over quality metrics describing high dimensional data.
%%
Their research agenda acts as guideline for our design, comprising applications using domain-experts, perceptual tuning towards human pattern recognition of important aspects and scalability between different data set sizes.
%%
%% TODO: Use this one? "Statistical resilience of random populations to random perturbations" - http://www.ncbi.nlm.nih.gov/pubmed/19256997 - wie eine Variable sich ändert, wenn man per Zufall die Populationen verändert, um zu schauen wie zuverlässig diese Statistik ist.
%% The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems
Ahmadi et al. \cite{Ahmadi} define the \emph{Sparse Regression Cube}, which partitions sparse high-dimensional into subspaces, which are then described using by their most reliable linear regression model.
%%
%In their application, they try to find most fuel-efficient roads connecting two user-defined landmarks.
%%
They focus on presenting a algebraic for efficient regression-model calculation to find the best fit for a subspace.
%%
The dimensions for the calculations are fixed, while we assess different features for each model calculation and can therefore not benefit from a hierarchical calculation approach.

%% Hierarchical Brushing of High-Dimensional Data Sets Using Quality Metrics
Albuquerque et al. \cite{Albuquerque} present an interactive exploration framework displaying quality metrics of high dimensional data sets with brushing facilities to create subsets.
%%
These subsets are then analyzed in detail, following a drilling-down approach.
%%
They incorporate scatter plot matrices (\texttt{SPLOM´S}) of quality metrics describing the features.
%%
We incorporate the idea of abstracting features for comparison.
%%
Turkay et al. \cite{Turkay} follow a similar approach by using both descriptive metrics for features as well as the features themselves and incorporate them into linked plots.
%%
The approach was applied to a cognitive aging studies, where new hypotheses could be derived.

@@

\paragraph{Prior Work.}
In recent publications, we analyzed the healthy aging process of the lumbar spine.
%%
Therefore, we utilized semiautomatic detection algorithm do detect the lumbar spine shape \cite{Klemm2013VMV}.
%%
The reliable part of the presented models are represented by the shape of the lumbar spine canal, which was extracted and described using line representation.
%%

We defined an Interactive Visual Analysis workflow for image-centric cohort study data, by extending the feature selection step (see \ref{sec:Background}) with an iterative analysis loop incorporating group selection and visualization using expert-input as well as clustering methods \cite{Klemm2014VIS}.
%%
Information visualizations of cohort study features were augmented with extracted image data, yielding linked views with both image-and non-image information.
%%
The popularity of the mosaic plot depicting \emph{Cram\'{e}r's V} contingency values among our domain experts was the inspiration for this work.
%%
The mosaic plot presented in \cite{Klemm2014VIS} only highlights pairwise correlation between features without a target disease.

%% TODO: Check DLPD for correct page numbers
While the presented techniques were well received with epidemiological experts, the explanatory power towards back pain was limited, yielding an analysis based on image-derived features, such as extracting and analyzing curvature, torsion and angle of the lumbar spine \cite{Klemm2015}.
%%
%These measures were analyzed using a novel decision tree quality plot towards non-image data.
%%
We concluded that the model quality is not enough to characterize back pain.
\\\\
The difference of the presented approach towards our previous and related work is twofold:
\begin{enumerate}
	\item We focus on large scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
	\item The analysis is conducted towards a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
\end{enumerate}
%%
\section{Regression Cube Analysis of Cohort Study Data}
% TODO: Do we need a preprocessing section?
%\com{Add Data Preprocessing Section.}
%%
The basic idea of our \emph{Regression Cube} is to provide an overview visualization of large cohort study data sets towards target variables.
%%
Overview visualizations of feature relationships as presented by Angelelli et al. \cite{Angelelli} are often focused on relationships between the visualized features.
%%
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
%%
In epidemiology, these relationships are also of interest, but rather w.r.t. their explanatory power towards the target feature.
%%
These target features often indicate the presence of the investigated disease.
%%
As described in Section \ref{sec:RegressionAnalysis}, regression analysis is the statistical tool of choice for analyzing these relationships.
%%
A regression model is based on expert knowledge, there is no unified rule on how to apply them to a given set of features, so they have to be applied with care.
%%
\subsection{Cube Description using Regression Formula Notation}
%%
Expert knowledge is introduced into regression analyses using the regression formulas.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%%
%In order to calculate all combinations of two to three independent features, we can apply  \emph{Regression Cube} aims to 
%%

Since we want to use the regression analyses with an overview visualization, we are are interested in all possible combinations of (two or more) independent features describing a target.
%%
We achieve this by introducing dynamic variables $X$, $Y$ and $Z$ into the regression notation.
%%
Our method replaces the dynamic variables with all features in the data set.
%%
In a data set with $n$ (e.g. 100) features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}
would yield $n^2$ (10.000) regression models, describing all possible combinations of two features in the data describing the target $Cancer$.
%%
The major advantage of this notation is that it comes natural to anyone familiar with regression analysis, because it uses the same notation. 
%%
%This allows for a fast adaptation and in the epidemiological application domain.
%%
With simple adjustments to the formula, different results can be achieved:
%%
\begin{itemize}
	\item $Z \sim X + Y$ calculates all combinations of two features towards all possible target features.
	\item $Cancer \sim X + Y + BodyWeight$ includes the $BodyWeight$ feature into all regression models as feature.
	\item $Cancer \sim X + Y + Z$ calculates all combinations of three features towards the target.
\end{itemize}
%%
The problem with this approach lies in its complexity.
%%
The number of calculated regression models increases exponentially for each dynamic variable added.
%%
If we assume a data set with 100 features with the formula $Z \sim X + Y$ we calculate 1,000,000 regression models.
%%
When each regression takes about 50~ms calculation, we have to wait roughly 14 hours for the calculation to complete.
%%
Therefore, the computational complexity needs to be reduced.
%%
%\com{Regression formulas are described using standard regression formula notation by introducing variables.}
%%
%\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}

\subsection{Target-Variable-dependent Dimension Reduction}
% \com{Uli, Can you please proofread this?}
% %%
% The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
% %%
% Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
% %%
% The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for this task \cite{CFS}.
% %%
% It uses information entropy to select the features which have the most explanatory power towards the target feature.
% %%
% At the same time it tries to reduce the correlation between the independent features.
% %%
In epidemiological studies, manifold recordings lead to an abundance of features.
%%
Generally many of them exhibit low or none correlation towards the target feature.
%%
Identifying irrelevant features and excluding them from the feature space considerably reduces computational costs and yields a comprehensible \emph{Regression Cube} representation.
%%
The Correlation-based feature selection (CFS) \cite{CFS} aims to find a feature subset that maximizes the \emph{merit value} $M_F$ which is the ratio between the average feature-class and feature-feature dependencies of features in F.
%%
The dependency of a set of features utilizes the entropy-based information gain to measure the explanatory power towards the target feature.
%%
Beginning with an empty set of features $F$ the CFS algorithm iteratively adds the feature $f$ to $F$ that leads to the highest new merit-value $M_{F\cup f}$ and halts when adding any feature does not increase the merit.
%%
When for example the \emph{body weight} has a strong explanatory power towards the target, it is likely that \emph{BMI} or \emph{waist circumference} exhibit similar correlations to the target.
%%
They, however, correlate strongly with each other.
%%
The CFS algorithm would then select the feature which has the largest explanatory power and discards the other dimensions.

@@

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
$R^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of $R^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the $R^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
Our goal is to create an overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

@@

 The values of the currently selected slice are mirrored and represented as orange data points on the slicing plane.

@@

%More importantly, it allows the user to focus all mental resources on the analysis tasks, rather than wrangling with configuring the system to serve the current task or hypothesis.

@@

Our prototype consists of three components:

@@

	\item The \emph{formula editor} allows to adjust the formula w.r.t. a hypothesis or conducting a \emph{hypothesis-free} analysis.

@@

\subsection{Regression Cube Visualization [2.SI].}
%%
The visualization and interaction with the \emph{Regression Cube} is the prototype core.
%%
Results from the statistical processors are uploaded into the visualization slice by slice, allowing the assessment of the data as soon as parts of the calculations are finished while the rest is still in processing.
